---
title: "Individual Assignment 5"
author: "Zelda Reif"
date: "2026-05-10"
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
error: FALSE
warning: FALSE
message: FALSE
---

# 1. Set Up

## Packages

```{r}
#| label: packages

# general use
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels
library(scales)

# reading .xlsx files 
library(readxl)

# visualize missing data
library(naniar)

# making plots
library(patchwork)
library(NatParksPalettes)
```

## Data

```{r}
#| label: data

# taxonomic info
taxon_list <- read_csv(here("data","taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects"
)

# site info
sites <- read_xlsx(
   here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
   sheet = "sites"
)

# water quality
water_quality <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Water Quality",
  na = c("", "n/a", "N/A", "over", "173+")
)
```

# 2. Cleaning

## `sites` object

```{r}
#| label: ncos-sites

# creating new object
ncos_sites <- sites |>
  # filtering to only include ncos quarterly sampling sites
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```

## clean `taxon_list`

```{r}
#| label: cleaning-taxon

taxon_list_clean <- taxon_list |> 
  # convert taxon names to snakecase
  mutate(verbatim_name = to_any_case(verbatim_name, 
                                     case = "snake"))
```

## clean `water_quality`

```{r}
#| label: cleaning-water-quality

water_quality_clean <- water_quality |> 
  clean_names() |> 
  semi_join(ncos_sites, by = "site") |> 
  unite("date_site", date, site, remove = FALSE) |> 
  group_by(date, site, date_site) |>
  summarize(med_ph = median(p_h, na.rm = TRUE),
             med_sal = median(salinity_ppt, na.rm = TRUE),
             med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  ungroup() |> 
  filter(date_site != "2022-08-12_NVBR")

```

## clean `aq_ins`

```{r}
#| label: cleaning-aq-ins

# creating new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # renaming columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter to only include "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert the data frame to long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance") |> 
  # group by date, site, taxon
  group_by(date, site, taxon) |> 
  # calculate average abundance per liter (sum abundance divided by 7.5 liters)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(
  select(water_quality_clean,
         date_site,
         med_ph,
         med_sal,
         med_do),
  by = "date_site"
) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # setting factor levels for site
  mutate(site_full = fct_reorder(site_full, med_sal, .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))
```

## clean `aq_ins_clean`
```{r}
#| label: cleaning-copepods

# creating new object from clean aquatic inverts 
copepods <- aq_ins_clean |> 
  # filter to only include copepods
  filter(taxon == "copepod") |> 
  # log + 1 transform mean abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1))

```

# Problems

## 1. Visualizing differences across sites in major taxon abundance

### a. Plan your figure

- x-axis: site
- y-axis: ostracod abundance
- geometry to represent average abundance/L: `geom_jitter`
- geometry to represent medians: `stat_summary()`

### b. Wrangle your data

```{r}
#| label: wrangling-aq-ins-clean

# create new object from aq_ins_clean
ost_abundance <- aq_ins_clean |> 
  # filter to only ostracods
  filter(taxon == "ostracod") |> 
  # remove missing abundance/L values
  filter(!is.na(ave_lit)) |> 
  # group by site
  group_by(site_full) |> 
  # log + 1 transform mean abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1)) |> 
  # ungroup data frame
  ungroup()

slice_sample(ost_abundance, n = 10)
```

### c. Code your figure

```{r}
#| label: plotting-ostracods
#| fig-height: 8
#| fig-width: 8

# base layer
ostracods_plot <- ggplot(data = ost_abundance,
                   # x-axis: site
                   # y-axis: mean abundance/L
                   mapping = aes(x = site_full,
                                 y = log_ave_lit)) +
  # first layer: points to represent observations
  geom_jitter(height = 0,
              shape = 21,
              width = 0.1) +
  # second layer: points to represent medians
  stat_summary(geom = "point",
               fun = "median",
               size = 4,
               color = "darkgreen") +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(15)) +
  # relabelling x- and y-axis, adding title
  labs(x = "Site",
       y = "log + 1 transformed average abundance/L",
       title = "Ostracod Abundance") +
  # different theme
  theme_bw()

# displaying object
ost_abundance
```

### d. Create a multipanel figure

```{r}
#| label: combining-plots
#| fig-height: 14
#| fig-width: 14

# base layer
salinity_plot <- ggplot(data = aq_ins_clean,
       # x-axis: site
       # y-axis: mean salinity
       mapping = aes(x = site_full,
                     y = med_sal)) +
  # first layer: salinity measurements at each survey
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(width = 15)) +
  # second layer: points to represent means
  stat_summary(fun = median,
               color = "red",
               size = 1) +
  # relabelling x- and y-axes
  labs(x = "Site",
       y = "Median salinity (ppt)",
       title = "Site Salinity") +
  # different theme
  theme_bw() 

# base layer
copepods_plot <- ggplot(data = copepods,
                   # x-axis: site
                   # y-axis: mean abundance/L
                   mapping = aes(x = site_full,
                                 y = log_ave_lit)) +
  # first layer: points to represent observations
  geom_jitter(height = 0,
              shape = 21,
              width = 0.1) +
  # second layer: points to represent medians
  stat_summary(geom = "point",
               fun = "median",
               size = 4,
               color = "blue") +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(15)) +
  # relabelling x- and y-axis, adding title
  labs(x = "Site",
       y = "Average abundance/L",
       title = "Copepod Abundance") +
  # different theme
  theme_bw() 

# base layer
ostracods_plot <- ggplot(data = ost_abundance,
                   # x-axis: site
                   # y-axis: mean abundance/L
                   mapping = aes(x = site_full,
                                 y = log_ave_lit)) +
  # first layer: points to represent observations
  geom_jitter(height = 0,
              shape = 21,
              width = 0.1) +
  # second layer: points to represent medians
  stat_summary(geom = "point",
               fun = "median",
               size = 4,
               color = "darkgreen") +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(15)) +
  # relabelling x- and y-axis, adding title
  labs(x = "Site",
       y = "Average abundance/L",
       title = "Ostracod Abundance") +
  # different theme
  theme_bw()

(salinity_plot) / (copepods_plot + ostracods_plot) +
  plot_annotation(title = "Comparing Salinity, Copepod Abundance, and Ostracod Abundance across sites at NCOS")
```

### e. Interpret your figure

South of Phelps Bridge and Devereux Creek had the highest Average abundance/L measurements across sites for Ostracods. The other sites had pretty consistent lower average abundance/L measurements. North of Venoco Bridge and South of Phelps Bridge had the highest Average abundance/L measurements across sites for Copepods. Main Channel and North of Phelps Bridge had the lowest average abundance/L measurements for Copepods. I could determine these comparisons by observing how high up on the y axis data points lie, and compare across sites on the x axis.

There does not appear to be a strong correlation between site salinity and ostracod abundance. The site with the highest average salinity (ppt) measurements, South of Vencoco Bridge, does not have a notably different average abundance/L measurements than the other sites. Similarly, the site with the highest average abundance/L measurements, South of Phelps Bridge, did not have comparibly high of low salinity (ppt) points along the y axis.

## 2. Visualizing total abundance as a function of salinity

### a. Plan your figure

x-axis: med_sal
y-axis: log_ave_lit
color: taxon
geometry: geom_point()
facets: facet_wrap(~ taxon, scales = "free")

### b. Wrangle your data

```{r}
#| label: cleaning-abu-sal

# create new object from aq_ins_clean
abundance_salinity <- aq_ins_clean |> 
  # create log-transformed abundance column
  mutate(log_ave_lit = log(ave_lit + 1)) |>
  # remove underscores from taxon names
  mutate(taxon = str_to_sentence(str_replace_all(taxon, "_", " "))) |>
  # group by taxon
  group_by(taxon) |>
  # calculate max abundance
  mutate(max_abundance = max(log_ave_lit, na.rm = TRUE)) |>
  # un group data frame
  ungroup() |>
  # reorder taxa by maximum abundance
  mutate(taxon = fct_reorder(taxon,
                        max_abundance,
                        .desc = TRUE)) |>
  # keep needed columns
  select(taxon,
         med_sal,
         log_ave_lit)

# display 10 random rows
abundance_salinity |>
  slice_sample(n = 10)
```

### c. Code your figure

```{r}
#| label: abu-sal-plot
#| fig-height: 10
#| fig-width: 10

# base layer
abu_sal_plot <- ggplot(data = abundance_salinity,
                   # x-axis: salinity
                   # y-axis: mean abundance/L
                   mapping = aes(x = med_sal,
                                 y = log_ave_lit,
                                 color = taxon)) +
  # first layer: points to represent observations
  geom_jitter(height = 0,
              shape = 21,
              width = 0.1) +
  # second layer: points to represent medians
  stat_summary(geom = "point",
               fun = "median",
               size = 1) +
  facet_wrap(~ taxon) + 
  # re labelling x- and y-axis, adding title
  labs(x = "Salinity (ppt)",
       y = "Average abundance/L",
       color = "Taxon",
       title = "Salinity vs. Abundance by Taxon at NCOS") +
  # change color palette, indicate 9 colors for 9 taxon groups
  scale_color_manual(values = natparks.pals("Everglades", n = 9)) +
  # different theme
  theme_bw() 

# display plot
abu_sal_plot
```

### d. Interpret your figure

In 3-5 sentences, describe differences or similarities between taxa in relationships between abundance and salinity. Be specific using the visual components of the plot.
