---
title: "Individual Assignment 5"
author: "Zelda Reif"
date: "2026-05-10"
format: pdf
---

# 1. Set Up

## Packages

```{r}
#| label: packages

# general use
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels
library(scales)

# reading .xlsx files 
library(readxl)

# visualize missing data
library(naniar)

# arrange plots
library(patchwork)
```

## Data

```{r}
#| label: data

# taxonomic info
taxon_list <- read_csv(here("data","taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects"
)

# site info
sites <- read_xlsx(
   here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
   sheet = "sites"
)

# water quality
water_quality <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Water Quality",
  na = c("", "n/a", "N/A", "over", "173+")
)
```

# 2. Cleaning

## `sites` object

Goal: create a new object from `sites` thst only includes NCOS sites


```{r}
#| label: ncos-sites

# creating new object
ncos_sites <- sites |>
  # filtering to only include ncos quarterly sampling sites
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```

## clean `taxon_list`

Goal: convert character strongs in existing `taxon_list` data frame into snakecase

```{r}
#| label: cleaing-taxon

taxon_list_clean <- taxon_list |> 
  # convert taxon names to snakecase
  mutate(verbatim_name = to_any_case(verbatim_name, 
                                     case = "snake"))
```

## clean `water_quality`

Goal: create a data frame with mean pH, mean DO (mg/L) and mean salinity (ppt) for each data and site

- create a new object from water_quality
- clean column names use janitor
- only include surveys from quarterly ncos sites
- create a new column called date_site to join with later - - data frames
- group bt data_site
- calc median pH, salinity and dissolved oxygen
- ungroup
- filter out outliers

```{r}
#| label: cleaning-water-quality

water_quality_clean <- water_quality |> 
  clean_names() |> 
  semi_join(ncos_sites, by = "site") |> 
  unite("date_site", date, site, remove = TRUE) |> 
  group_by(date, site) |> 
  summarize(med_ph = median(p_h, na.rm = TRUE),
             med_sal = median(salinity_ppt, na.rm = TRUE),
             med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE) |> 
  ungroup() |> 
  filter(date_site != "2022-08-12_NVBR")

```


## clean `aq_ins`

Goal: create long format data frame in which each row contains the counted individuals in each taxon during each survey. The data frame should include taxonomic information foe each organism and the water quality info for each survey.

- create new object
- clean column names
- filter to only include a quarterly NCOS sites
- slect columns
- filter to only include "filtered beaker" smaple
- convert data frame to long format
- group by site, date and taxon
- calculate avg. abundance per liter (sum counts divided by 7.5 liter)
- create new column called date_site to join with water
- quality
- join data frame
- join with taxon list

```{r}
#| label: cleaning-aq-ins

aq_ins_clean <- aq_ins |> 
  clean_names() |> 
  semi_join(ncos_sites, by = "site") |> 
  rename(date = date_on_vial) |> 
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         nematode,
         diptera_ceratopogonidae,
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  filter(sample_type %in% c("FB250", "FB 250")) |> 
  pivot_longer(
    # columns to make longer
    cols = ostracod:annelida_polychaete,
    # new column for the old column names
    names_to = "taxon",
    # new column for the values in each of the old columns
    values_to = "abundance"
  ) |> 
  group_by(date, site, taxon) |> 
  summarize(ave_lit = sumabundance, na.rm = TRUE/7.5) |>
  ungroup() |> 
  unite("date_site", date, site, remove = TRUE) |> 
  left_join(water_quality_clean, by = "date_site") |> 
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  mutate(site_full = case_when)


```

# Problems

## 1. 