---
title: "Individual Assignment 5"
author: "Zelda Reif"
date: "2026-05-10"
format: pdf
error: FALSE
message: FALSE
---

# 1. Set Up

## Packages

```{r}
#| label: packages

# general use
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels
library(scales)

# reading .xlsx files 
library(readxl)

# visualize missing data
library(naniar)

# arrange plots
library(patchwork)
```

## Data

```{r}
#| label: data

# taxonomic info
taxon_list <- read_csv(here("data","taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects"
)

# site info
sites <- read_xlsx(
   here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
   sheet = "sites"
)

# water quality
water_quality <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Water Quality",
  na = c("", "n/a", "N/A", "over", "173+")
)
```

# 2. Cleaning

## `sites` object

Goal: create a new object from `sites` thst only includes NCOS sites


```{r}
#| label: ncos-sites

# creating new object
ncos_sites <- sites |>
  # filtering to only include ncos quarterly sampling sites
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```

## clean `taxon_list`

Goal: convert character strongs in existing `taxon_list` data frame into snakecase

```{r}
#| label: cleaing-taxon

taxon_list_clean <- taxon_list |> 
  # convert taxon names to snakecase
  mutate(verbatim_name = to_any_case(verbatim_name, 
                                     case = "snake"))
```

## clean `water_quality`

Goal: create a data frame with mean pH, mean DO (mg/L) and mean salinity (ppt) for each data and site

- create a new object from water_quality
- clean column names use janitor
- only include surveys from quarterly ncos sites
- create a new column called date_site to join with later - - data frames
- group bt data_site
- calc median pH, salinity and dissolved oxygen
- ungroup
- filter out outliers

```{r}
#| label: cleaning-water-quality

water_quality_clean <- water_quality |> 
  clean_names() |> 
  semi_join(ncos_sites, by = "site") |> 
  unite("date_site", date, site, remove = FALSE) |> 
  group_by(date, site, date_site) |>
  summarize(med_ph = median(p_h, na.rm = TRUE),
             med_sal = median(salinity_ppt, na.rm = TRUE),
             med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  ungroup() |> 
  filter(date_site != "2022-08-12_NVBR")

```


## clean `aq_ins`

Goal: create long format data frame in which each row contains the counted individuals in each taxon during each survey. The data frame should include taxonomic information foe each organism and the water quality info for each survey.

- create new object
- clean column names
- filter to only include a quarterly NCOS sites
- slect columns
- filter to only include "filtered beaker" smaple
- convert data frame to long format
- group by site, date and taxon
- calculate avg. abundance per liter (sum counts divided by 7.5 liter)
- create new column called date_site to join with water
- quality
- join data frame
- join with taxon list

```{r}
#| label: cleaning-aq-ins

# creating new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # renaming columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter to only include "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert the data frame to long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance") |> 
  # group by date, site, taxon
  group_by(date, site, taxon) |> 
  # calculate average abundance per liter (sum abundance divided by 7.5 liters)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(
  select(water_quality_clean,
         date_site,
         med_ph,
         med_sal,
         med_do),
  by = "date_site"
) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # setting factor levels for site
  mutate(site_full = fct_reorder(site_full, med_sal, .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))
```

# Problems

## 1. 

```{r}

```

