---
title: "Individual Assignment 5"
author: "Zelda Reif"
date: "2026-05-10"
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
error: FALSE
warning: FALSE
message: FALSE
---

# 1. Set Up

## Packages

```{r}
#| label: packages

# general use
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels
library(scales)

# reading .xlsx files 
library(readxl)

# visualize missing data
library(naniar)

# arrange plots
library(patchwork)
```

## Data

```{r}
#| label: data

# taxonomic info
taxon_list <- read_csv(here("data","taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects"
)

# site info
sites <- read_xlsx(
   here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
   sheet = "sites"
)

# water quality
water_quality <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Water Quality",
  na = c("", "n/a", "N/A", "over", "173+")
)
```

# 2. Cleaning

## `sites` object

```{r}
#| label: ncos-sites

# creating new object
ncos_sites <- sites |>
  # filtering to only include ncos quarterly sampling sites
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```

## clean `taxon_list`

```{r}
#| label: cleaing-taxon

taxon_list_clean <- taxon_list |> 
  # convert taxon names to snakecase
  mutate(verbatim_name = to_any_case(verbatim_name, 
                                     case = "snake"))
```

## clean `water_quality`

```{r}
#| label: cleaning-water-quality

water_quality_clean <- water_quality |> 
  clean_names() |> 
  semi_join(ncos_sites, by = "site") |> 
  unite("date_site", date, site, remove = FALSE) |> 
  group_by(date, site, date_site) |>
  summarize(med_ph = median(p_h, na.rm = TRUE),
             med_sal = median(salinity_ppt, na.rm = TRUE),
             med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  ungroup() |> 
  filter(date_site != "2022-08-12_NVBR")

```

## clean `aq_ins`

```{r}
#| label: cleaning-aq-ins

# creating new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # renaming columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter to only include "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert the data frame to long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance") |> 
  # group by date, site, taxon
  group_by(date, site, taxon) |> 
  # calculate average abundance per liter (sum abundance divided by 7.5 liters)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(
  select(water_quality_clean,
         date_site,
         med_ph,
         med_sal,
         med_do),
  by = "date_site"
) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # setting factor levels for site
  mutate(site_full = fct_reorder(site_full, med_sal, .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))
```

# Problems

## 1. Visualizing differences across sites in major taxon abundance

### a. Plan your figure

- x-axis: site
- y-axis: ostracod abundance
- geometry to represent average abundance/L: `geom_jitter`
- geometry to represent medians: `stat_summary()`

### b. Wrangle your data

```{r}
#| label: wrangling-aq-ins-clean

# create new object from aq_ins_clean
ost_cod_abundance <- aq_ins_clean |> 
  # filter to only ostracods
  filter(taxon == "ostracod") |> 
  # remove missing abundance/L values
  filter(!is.na(ave_lit)) |> 
  # group by site
  group_by(site_full) |> 
  # combines all abundance values for each site
  summarize(
    ave_lit = sum(ave_lit, na.rm = TRUE)) |> 
  # ungroup dataframe
  ungroup()

slice_sample(ost_cod_abundance, n = 10)
```

### c. Code your figure

```{r}

```

### d. Create a multipanel figure

```{r}

```

### e. Interpret your figure

differences between sites in copepod and ostracod abundance
how ostracod abundance may relate (or not) to site-level differences in salinity

## 2. Visualizing total abundance as a function of salinity

### a. Plan your figure

x-axis:
y-axis:
color:
geometry:
facets:

### b. Wrangle your data

```{r}

```

### c. Code your figure

```{r}

```

### d. Interpret your figure

In 3-5 sentences, describe differences or similarities between taxa in relationships between abundance and salinity. Be specific using the visual components of the plot.
